Variable Density Tesseroids: Gravity fields calculation in spherical coordinates using variable densities in depth

### 2017-10-25 manuscript/manuscript.tex
Modified manuscript

In order to make the computation more efficient, we use only second order GLQ for each integration while we ensure the accuracy of the method through the application of the modified adaptive discretization algorithm developed by \citet{Uieda2016}.
The modifications made by \citet{Uieda2016} to the original adaptive discretization method \citep{Li2011} can be summed up in:
(1) a faster calculation of the distance from the computation point to the tesseroid and 
(2) the application of a stack based algorithm that speeds up the computation and gives more control over the recursion step.

More specifically, \citet{Uieda2016} redefined the dimensions of an arbitrary tesseroid: $L_\lambda$, $L_\phi$ and $L_r$. The former ones are the arc-distances measured along its top surface while $L_r$ is the difference between the radii of the top and bottom surfaces.

The adaptive discretization algorithm starts by checking that the following inequality is satisfied:

@@

\noindent where $i \in \{\lambda, \phi, r\}$, $d$ is the distance between the geometric centre of the tesseroid and the computation point, and $D$ is a predefined value called distance-size ratio.
If the inequality \ref{eq:distance-size-ratio} is not held for any spherical coordinate, then the tesseroid must be divided by half in that direction.
A more complete description of the modified adaptive discretization algorithm can be found in \citet{Uieda2016}.
On the contrary, if the inequality is satisfied, the algorithm computes the desired gravity field using the GLQ approximation.

In summary, the distance-size ratio $D$ determines indirectly how many times the tesseroids will be divided, and therefore regulates both the accuracy of the algorithm and its computation time.
In this way the modified version of the adaptive discretization algorithm efficiently improves its accuracy and gives the user more control over the computation: it allows to stop the division of the tesseroids when their dimensions are lower than a certain value (e.g. less than 10 cm), in which case it does not modifies significantly the desired gravitational field.

@@

This algorithm can be subsequently applied to every subdividing tesseroid, until the inequation \ref{eq:delta-density} is not held or the subdivisions are smaller than 1mm.
This two conditions guarantee that the algorithm is finite, i.e. it does not fall into an infinite loop. \todo{chequear esta oracion}
Once this conditions are met, every subdividing tesseroid is subjected to the modified adaptive discretization algorithm and the GLQ in order to compute the gravity field generated by it.

It's easy to see that the delta ratio indirectly controls how many times the tesseroids will be subdivided based on the variation of the density function. Therefore it controls the accuracy involving the numerical error that the density variation introduces and also the computation time.
This raises the need to determine a maximum value of $\delta$ that ensures an acceptable accuracy while minimising the computation time.
We will perform this determination with the same methodology as the one applied to the $D$ ratio: compare the numerical model with the analytical solutions for a spherical shell.

@@

In order to determine the minimum value for the distance-size ratio $D$ such the numerical approximation achieves an acceptable accuracy we have compared the numerical model with the analytical solutions for a spherical shell with variable density for three special cases: a linear, an exponential and a discontinuous density function.
The first one constitutes the simplest variation, while the second one represents a higher varying function. The later one tries to emulate a contact between two different media (e.g. crust and mantle interface).

@@

We computed these gravity fields on every point of the four grids for different values of $D$, ranging from 0.5 to 10 with a step of 0.5.
Then we calculated the maximum difference between these results and the value that the analytical solution assumes at the same height.

@@

Moreover, we did not apply the density-based discretization algorithm in these tests in order to isolate the behaviour of the adaptive discretization algorithm.

@@

\subsection{Linear Density}
Firstly,  we proposed a spherical shell with linear density in such way that it assumes 2670~kg/m$^3$ on the surface of the Earth and 3300~kg/m$^3$ at the inner radius:

@@

In Figures \ref{fig:D-linear-thin} and \ref{fig:D-linear-thick} we can see the results for the potential, the vertical component of the gradient $g_z$ and the $g_{zz}$ component of the Marussi tensor.
Because the other two diagonal components of the tensor show similar results as the ones obtained for the $g_{zz}$, we exclude them from the figure, although they can be found in the repository. \todo{chequear}

@@

It's worth to remember that the density-based discretization algorithm is not applied for linear density function comparisons, thus the adaptive discretization algorithm is the only mechanism that controls the accuracy of the approximation.

@@

\noindent and the value of $b$ equal to the thickness of the spherical shell we use to perform the test.
In this way, the density function assumes the values of 2670 kg/m$^3$ and 3300 kg/m$^3$ on the shell's outer and inner radii, respectively.
The analytical solution of the gravity potential is easily obtained as the sum the solution for a pure exponential function and the one for an homogeneous spherical shell.

@@

Because the constant $b$ controls the variation of the density function, we wanted to test how does the accuracy of the method behaves at different values of $b$ when the density-based discretization algorithm is not applied.
In order to do this we proposed a set of density functions, each one with a different value of $b$.
We computed the difference between the Marussi tensor component $g_{zz}$ generated by the numerical model on the big grid of 30$^\circ\times$30$^\circ$ (see Table \ref{tab:grids}) and the corresponding analytical solution.
These differences were calculated for each proposed density function and for the same set of values of $D$ we have been using on the previous tests, obtaining one curve for each value of $b$.

>>>>

The main differences between the modified version made by \citet{Uieda2016} and the original one developed by \citet{Li2011} can be summed up in:
(1) a faster calculation of the distance from the computation point to the tesseroid and 
(2) the application of a stack based algorithm that speeds up the computation and gives more control over the recursion step.
More specifically, \citet{Uieda2016} redefined the dimensions of an arbitrary tesseroid: $L_\lambda$, $L_\phi$ and $L_r$. The former ones are the arc-distances measured along its top surface while $L_r$ is the difference between the radii of the top and bottom surfaces.

While keeping the GLQ fixed on second orders, the adaptive discretization algorithm starts by checking that the following inequality is satisfied:

@@

\noindent where $i \in \{\lambda, \phi, r\}$, $d$ is the distance between the geometric centre of the tesseroid and the computation point, and $D$ is a predefined value called distance-size ratio.
If the inequality \ref{eq:distance-size-ratio} is not held for at least one spherical coordinate, then the tesseroid must be divided by half in that direction or directions.
A more complete description of the modified adaptive discretization algorithm can be found in \citet{Uieda2016}.
On the contrary, if the inequality is satisfied, the algorithm computes the desired gravity field using the GLQ approximation.

In summary, the distance-size ratio $D$ determines indirectly how many times the tesseroids will be divided, and therefore efficiently regulates both the accuracy of the algorithm and its computation time.
Moreover, it limits the computation time by stopping the division of the tesseroids when their dimensions are lower than a certain value (1mm), assuming that such small tesseroids do not modifies significantly the desired gravitational field.

@@

This algorithm can be subsequently applied to every subdividing tesseroid, until the inequation \ref{eq:delta-density} is not held or the subdivisions are smaller than 1mm.
This two conditions guarantee that the algorithm is finite.
Once this conditions are met, every subdividing tesseroid is subjected to the modified adaptive discretization algorithm and the GLQ in order to compute the gravity field generated by it.

It's easy to see that the delta ratio indirectly controls how many times the tesseroids will be subdivided based on the variation of the density function.
Therefore it determines the accuracy involving the numerical error that the density variation introduces and also the computation time.
This raises the need to determine a maximum value of $\delta$ that ensures an acceptable accuracy while minimising the computation time.
We will perform this determination with the same methodology as the one applied to the $D$ ratio: compare the numerical model with the analytical solutions for a spherical shell.

@@

In order to determine the minimum value for the distance-size ratio $D$ such the numerical approximation achieves an acceptable accuracy we have compared the numerical model with the analytical solutions for a spherical shell with variable density for three special cases: a linear, an exponential and a discontinuous density function.
The first one constitutes the simplest variation, the second one represents a higher varying function, while the later one tries to emulate a contact between two different media (e.g. crust and mantle interface).

@@

We computed these gravity fields on every point of the four grids for different values of $D$, ranging from 0.5 to 10 with a step of 0.5, without applying the density-based discretization algorithm in order to isolate the behaviour of the adaptive discretization one.
Then we calculated the maximum difference between these results and the value that the analytical solution assumes at the same height.

@@

@@

\subsection{Linear Density}
Firstly,  we proposed a spherical shell with linear density in such way it assumes 2670~kg/m$^3$ on the surface of the Earth and 3300~kg/m$^3$ at the inner radius:

@@

We can see the results for the potential, the vertical component of the gradient $g_z$ and the $g_{zz}$ component of the Marussi tensor on Figures \ref{fig:D-linear-thin} and \ref{fig:D-linear-thick}.
We exclude the results corresponding to the other two diagonal components of the tensor because they are similar to the ones obtained for the $g_{zz}$, although they can be found on the repository. \todo{chequear}

@@

@@

\noindent and the value of $b$ equal to the thickness of the spherical shell we use to perform the test.
In this way, the density function assumes the values of 2670 kg/m$^3$ and 3300 kg/m$^3$ on the shell's outer and inner radii, respectively.
The analytical solution of the gravity potential is easily obtained as the sum of the ones for a pure exponential function and for an homogeneous spherical shell.

@@

Moreover, we can see that the differences for the potential and the $g_z$ component of the gradient keep constant even for high values of $D$.
Our interpretation is that the modified adaptive discretization algorithm is limited when increasing the accuracy for this density function, supporting the need of a density-based discretization algorithm.

Because the constant $b$ controls the variation of the density function, we wanted to test how does the accuracy of the method behaves at different values of $b$ when the density-based discretization algorithm is not applied.
In order to do this we proposed a thin spherical shell (1km of thickness) with a set of density functions, each one with a different value of $b$.
We computed the difference between the Marussi tensor component $g_{zz}$ generated by the numerical model on the big grid of 30$^\circ\times$30$^\circ$ (see Table \ref{tab:grids}) and the corresponding analytical solution.
These differences were calculated for each proposed density function and for the same set of values of $D$ we have been using on the previous tests, obtaining one curve for each value of $b$.
